Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}

### 2017-12-22 drafts/elife/draft.tex
figure captions

\begin{abstract}
  Recent advances in neural recording technologies have enabled
  simultaneous measurements of the majority of head ganglia neurons in
  both immobilized and freely-behaving
  \celegans~\citep{schrodel2013brain, prevedel2014simultaneous,
    nguyen2016whole}.  The dynamics of neural activity shed light on
  how \celegans~processes sensory information and generates motor
  activity.  To understand these dynamics, we develop recurrent state
  space models that decompose complex time-series into segments with
  simple, linear dynamics. We incorporate these models into a robust,
  hierarchical framework for combining information across whole-brain
  recordings of many worms.  Using this framework, we reveal latent
  states of population neural activity, along with the discrete
  behavioral modes that drive dynamics in this latent state space.  We
  find stochastic transition patterns between these discrete
  behavioral modes, and we see that transition probabilities are
  determined by a combination of current brain state and environmental
  cues.  In addition to quantifying neural dynamics, this
  probabilistic framework aids in neural identification---currently a
  laborious, manual task---and reveals clusters of neurons that are
  similarly tuned in latent state space.  Finally, we find a
  significant overlap between our inferred modes and the
  manually-labeled modes of~\citet{kato2015global}
  and~\citet{nichols2017global}, which were shown to correspond to
  different behaviors, like forward crawling, reversals, and
  turns. Our methods automatically discover and quantify these
  behaviorally-meaningful states directly from neural activity,
  yielding powerful new tools for neuro-behavioral analysis.
\end{abstract}

@@

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/figure1} 
\caption{
  % A schematic of the hierarchical model for combining
  % information across multiple worms in order to learn a shared
  % dynamical system for \celegans.  \textbf{a.}~All worms share a
  % canonical dynamics library. Here, we have three dynamics functions
  % (blue, red, gold), each of which is a simple linear mapping from
  % $x_t \in \reals^2$ to $x_{t+1} \in \reals^2$.  The recurrent model
  % assigns probabilities to the different dynamics functions based on
  % the current location in $x$-space.  \textbf{b.}~The worms also share
  % an emission matrix, $C$, which can be viewed as an embedding of
  % neurons in latent space.  \textbf{c.}~Each worm has its own dynamics
  % library, which allows for slight deviations from the canonical
  % dynamics.  \textbf{d.}~The model segments neural activity into a
  % sequence of discrete states~$z_{t} \in \{1, 2, 3\}$, which index
  % into the dynamics library.  \textbf{e.}~The discrete states specify
  % the dynamics that are used to propagate the continuous latent
  % states~$x_t$ forward in time. Here, the two dimensional latent
  % states are colored according to the instantaneous discrete
  % state. \textbf{f.}~The observed neural activity is a linear
  % projection of the underlying continuous latent states.  On any given
  % trial, we observe only a fraction of all the neurons.  The
  % hierarchical model combines these partial recordings to infer the
  % underlying states, the dynamics libraries, and the shared neural
  % embedding.
}
\label{fig:model}
\end{figure}

@@

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/figure2} 
\caption{}
\label{fig:emissions}
\end{figure}

@@

\begin{figure}[t]
\centering%
\includegraphics[width=6in]{figures/figure3} 
\caption{}
\label{fig:syllables}
\end{figure}

@@

\begin{figure}[t]
\centering%
\includegraphics[width=6in]{figures/figure4} 
\caption{}
\label{fig:recurrent}
\end{figure}

@@

\begin{figure}[t]
\centering%
\includegraphics[width=6in]{figures/figure5} 
\caption{}
\label{fig:o2}
\end{figure}

Having demonstrated the utility of

>>>>

\begin{abstract}
  Recent advances in neural recording technologies have enabled
  simultaneous measurements of the majority of head ganglia neurons in
  both immobilized and freely-behaving
  \celegans~\citep{schrodel2013brain, prevedel2014simultaneous,
    nguyen2016whole}.  The dynamics of neural activity shed light on
  how \celegans~processes sensory information and generates motor
  activity.  To understand these dynamics, we develop recurrent state
  space models that decompose complex time-series into segments with
  simple, linear dynamics. We incorporate these models into a robust,
  hierarchical framework for combining information across whole-brain
  recordings of many worms.  Using this framework, we reveal latent
  states of population neural activity, along with the discrete
  behavioral modes that drive dynamics in this latent state space.  We
  find stochastic transition patterns between these discrete
  behavioral modes, and we see that transition probabilities are
  determined by a combination of current brain state and environmental
  cues.  In addition to quantifying neural dynamics, this
  probabilistic framework aids in neural identification---currently a
  laborious, manual task---and reveals clusters of neurons that are
  similarly tuned in latent state space.  Finally, we find a
  significant overlap between our inferred states and the
  manually-labeled states of~\citet{kato2015global}
  and~\citet{nichols2017global}, which were shown to correspond to
  different behaviors, like forward crawling, reversals, and
  turns. Our methods automatically discover and quantify these
  behaviorally-meaningful states directly from neural activity,
  yielding powerful new tools for neuro-behavioral analysis.
\end{abstract}

@@

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/figure1} 
\caption{
  \textit{A probabilistic framework for combining
  information across multiple worms in order to learn a shared
  dynamical system for \celegans.}
  \textbf{a.}~The data consists of observed neural activity collected
  from many different worms.  In each recording we observe a different
  subset of labeled neurons in the head ganglia.
  \textbf{b.}~We model the observed activity as a linear function of
  an underlying, continuous, and typically low-dimensional latent state,
  here shown as a two-dimensional signal for each recording.
  \textbf{c.}~The dynamics of these continuous states are governed by
  a co-evolving discrete state.
  \textbf{d.}~These discrete states index into a library of linear
  dynamics functions for each worm, here shown as maps from~$\reals^2$ to~$\reals^2$.
  Each dynamics function is selectively used in different regions of
  continuous latent space, as shown by the different intensities of
  the vector field.
  \textbf{e.}~To increase robustness to model misspecification, we
  use a multivariate-t distribution for the noise in the continuous
  state updates.
  \textbf{f.}~Each worm has its own dynamics library, but they are
  constrained to be close to the shared, canonical library.  This
  combines information to learn shared structure while also allowing
  for worm-to-worm variability.
  \textbf{g.} Likewise, we allow each worm to have its own observation
  variances to account for individual differences that may arise from,
  for example, variable levels of fluorescent protein expression.
  Missing neurons are denoted by gray boxes. 
}
\label{fig:model}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/figure2} 
\caption{ \textit{The mapping from continuous latent states to
    observations aids in neural identification and reveals clusters of
    neurons.}
  \textbf{a.}~The observed neural activity (solid black lines) in the
  first worm of the \citet{kato2015global} dataset along with the
  smoothed activity given by the model (blue lines). We use
  correlations learned across the population to predict the activity
  of unobserved neurons (those with dotted black lines).
  \textbf{b.}~These predictions are a useful aid for neural identification.
  We withheld the labels of a fraction of neurons and measured the similarity
  between their activity and the models predictions.  The similarity was
  significantly higher for the correct assignments as compared to incorrect
  assignments.
  \textbf{c.}~The probabilistic model offers a principled means of assessing
  the dimensionality of neural activity, even though only a subset of neurons
  are labeled on any given trial.  We find that ten latent dimensions yields
  the highest test likelihood, and that allowing per-worm observation variances
  yields substantial improvements in test performance.
  \textbf{d.}~The per-worm variances show that observation noise can
  vary substantially from one worm to another. Gray squares denote missing
  neurons.
  \textbf{e.}~We can view the linear mapping from latent states to observations
  as either a matrix or as an embedding of neurons.
  \textbf{f.}~Viewed as an embedding, we see that neurons cluster into
  correlated groups with similar tuning to the continuous latent state.
  \textbf{g.}~These groups roughly correspond to neurons at different
  locations along the worm's body.  }
\label{fig:emissions}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/figure3} 
\caption{ \textit{The hierarchical SLDS decomposes continuous states
    into segments with simple linear dynamics.}
  \textbf{a.}~Three dimensions of the inferred continuous latent states
  for the first worm, color coded by the corresponding discrete latent
  state.
  \textbf{b.}~We select the number of discrete states and justify the
  recurrent and robust extensions by measuring the marginal likelihood
  of held-out test data. We find that the robust, recurrent model
  with eight states performs best.
  \textbf{c.}~The inferred segmentations align with the manual segmentations
  of~\citet{kato2015global}.
  \textbf{d.}~Moreover, the inferred states are in close correspodence with
  the manually labeled states, which were shown to indicate different behaviors.
  \textbf{e.}~Each discrete state is associated with a canonical set of linear
  dynamics, which can be viewed as mappings from~$\reals^10$ to~$\reals^10$.
  We show the mappings for the first three dimensions here, with vectors placed
  at locations where each state was used.  We see a variety of rotational
  states that can be combined to yield dynamics like those in~\textbf{a.}.
  \textbf{f.}~We simulate these linear dynamics from a variety of starting
  locations for the fourth (green) and highlight four for closer inspection.
  \textbf{g.}~These four trajectories give rise to different activations of
  the neural clusters. We see see a dip then rise in cluster 3, which consists
  of neurons related to ventral turns, consistent with the manual labeling
  of the green state in~\textbf{d}.
}
\label{fig:syllables}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/figure4} 
\caption{ \textit{Discrete state transitions show a non-Markovian,
    chain-like structure.}  \textbf{a.}~Counts of inferred transitions
  from one state to another. The states are permuted to emphasize the
  feed-forward structure.  \textbf{b.}~An alternative view of the
  transitions; each state is a node in a graph and edges denote
  high-probability transitions.  This shows transitions between
  forward and reverse crawling linked by dorsal and ventral turns.
  \textbf{c.}~The standard SLDS assumes Markovian transitions, which
  imply a geometric distribution of state durations (dotted black
  lines), but Fig.~\ref{fig:syllables} clearly demonstrates that state
  transitions occur near the origin.  The recurrent SLDS incorporates
  this dependency and generates state durations (solid black lines)
  that are a much better fit to those inferred from the data (colored
  bars).  }
\label{fig:recurrent}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/figure5} 
\caption{\textit{Incorporating exogenous factors like oxygen level
    into the dynamics model yield differential responses in worms from
    different genetic strains and developmental stages.}
  \textbf{a.}~The observed neurons and their inferred observation variances
  for each of the 41 worms in the~\citet{nichols2017global} dataset.
  \textbf{b.}~Worms of different genetic strains (N2, npr1) and developmental
  stages (pre-lethargus, lethargus) exhibit different levels of overall
  state usage, with blue and yellow much more likely in the lethargus stage.
  \textbf{c.}~As above, the inferred states closely match the manually
  labeled states of~\citet{nichols2017global}, with blue and yellow mapping
  onto the forward (\textsf{Fwd}) and quiescent (\textsf{Q}) states.
  \textbf{d.}~We extended the SLDS to allow exogenous factors like oxygen
  level influence discrete transition probabilities.
  \textbf{e., f.}~Upon fitting the model, we find that oxygen has a limited
  effect on N2 worms and a much stronger effect on npr1 worms. In npr1 worms,
  increasing the oxygen level decreases the probability of quiescent and
  forward states and elicits reversals and turns.
}
\label{fig:o2}
\end{figure}

The preceding results demonstrated the efficacy of hierarchical,
recurrent, and robust SLDS for learning interpretable dynamics of
neural activity from the noisy, partial recordings
of~\citet{kato2015global}.  We now turn to the dataset studied
in~\citet{nichols2017global}, which contains many more worms with
roughly the same number of neurons per recording but is complicated by
additional factors.  First, there are two strains of worms (N2 and
npr-1) at two different stages of development (pre-lethargus and
lethargus); second, the oxygen level is manipulated over the course of
the recordings to elicit different responses.  Here we show how our
framework can be extended to study the effects of these manipulations.

Fig.~\ref{fig:o2}a shows which neurons were labeled in each of the
recordings along with the inferred observation variance. There are 41
worms, roughly ten from each strain and developmental stage
combination (N2 + pre-lethargus, N2 + lethargus, npr1 + pre-lethargus,
and npr1 + lethargus). Some neurons are consistently identified in
almost all worms; others are identified in only a handful.  As before,
our hierarchical framework combines all this information.

Upon fitting our model, we immediately see a difference in the overall
usage of the eight states (Fig.~\ref{fig:o2}b). The worms in the
lethargus stage predominantly use the blue and yellow states.  Again,
we found a close correspondence between our inferred states and the
manually labeled states of~\citet{nichols2017global}
(Fig.~\ref{fig:o2}c). Here, blue and yellow both map strongly onto the
quiescent (\textsf{Q}) state. This is consistent with the
characterization of the lethargus stage---at this stage, worms exhibit
much less crawling behavior. The pre-lethargus worms execute a broader
variety of behaviors, including reversals (\textsf{Rev.}) and dorsal
and ventral turns  (\textsf{DT} and \textsf{VT}, respectively).

A key finding of~\citet{nichols2017global} was that increasing the
oxygen level can elicit activity in lethargus npr1 worms but not in
lethargus N2 worms. We can quantify this effect with our models by
including oxygen level as an exogenous input.  Fig.~\ref{fig:o2}d
shows how the graphical model of the recurrent SLDS changes: now the
discrete state at time~$t$ (red circle) depends on the preceding discrete
state, the preceding continuous state, and the current oxygen level.
\todo[inline]{consider alternative models in which oxygen level
  influences continuous states and observations?}

Fig.~\ref{fig:o2}e and~\ref{fig:o2}f show the inferred effect of
oxygen level for each of the four strain/developmental stage pairs.
As predicted, only the npr1 worms show a much larger response to
oxygen. Increasing the oxygen level increases the probability
of reversals and turning related states (red, orange, purple) and
decreases the probability of forward and quiescent (blue, yellow)
states. Somewhat surprisingly, the magnitude of the effect is
comparable for both pre-lethargus and lethargus npr1 worms, even
though the baseline level of reversals and turns is much higher in
the pre-lethargus worms.
